---
title: "Cost effectiveness analysis of sertraline versus Amitriptyline"
subtitle: "markov cohort simulation with probability sensitivity analysis"
author: "Joshua Edefo"
date: "2024-01-11"
email: "edefojoshua2000@yahoo.com"
output: github_document
---
libraries
```{r a, message=FALSE}
# Markov cohort simulation with probability Sensitivity analysis of Sertraline
library(markovchain)
library(cartography)
library(rmarkdown)
library(sf)
library(sp)
```
Sertraline
```
Diagrammatic presentation of transition matrix
```{r ab}
Set.matrix<-new("markovchain", states=c("Depressed", "Borderline", "Depression_free"), 
         transitionMatrix=matrix(data=c(0.1, 0.3, 0.6, 0.3, 0.1, 0.6, 0.1, 0.1, 0.8), 
                                   byrow = TRUE, nrow =  3), name = "Sertraline") 
plot(Set.matrix, package = "daigram")
```
```{r b}
# R code inside a function
N_psa <- 10000

# distribution of transition probability matrix using beta distribution with SD= 25% of the mean value
params<-data.frame(
  p_Depressed_Depressed   = rbeta(N_psa, 10, 90),
  p_Depressed_Borderline  = rbeta(N_psa,  30, 70),
  p_Depressed_Depressionfree  = rbeta(N_psa,  60, 40),
  p_Bordeline_Depressed  = rbeta(N_psa,  30, 70),
  p_Borderline_Borderline = rbeta(N_psa, 10, 90),
  p_Borderline_Depressionfree = rbeta(N_psa,  60, 40),
  p_Depressionfree_Depressed  = rbeta(N_psa,  10, 90),
  p_Depressionfree_Borderline = rbeta(N_psa,  10,  90),
  p_Depressionfree_Depressionfree = rbeta(N_psa, 80, 20),
  
  # distribution of cost using gamma distribution with SD= 25% of the mean value
  c_Depressed = rgamma(N_psa, shape =16, scale = 1432/16),
  c_Borderline = rgamma(N_psa, shape =16, scale =4296/16),
  c_Depressionfree = rgamma(N_psa, shape =16, scale =8593/16),
  
  # distribution of effect using gamma distribution with SD= 25% of the mean value
  q_Depressed = rgamma (N_psa, shape =16, scale = 2/16),
  q_Borderline = rgamma(N_psa, shape =16, scale = 6/16),
  q_Depressionfree = rgamma(N_psa, shape =16, scale = 9/16)
)

model.se<- function(.params) { 
  with(.params, {
    
    n_t<-40     # no of cycles
    n_s<-3      # no of states
    n_c<-1000   #  no of persons in cohort
    
    v_state_names<-c("Depressed", "Borderline", "Depressionfree") # names of states
    
    # probability tranistion matrix
    m_P<-matrix(0, nrow = 3, ncol =3, 
                dimnames = list(from = v_state_names, to = v_state_names ))
    m_P["Depressed", "Depressed"]<-p_Depressed_Depressed 
    m_P["Depressed", "Borderline"]<- p_Depressed_Borderline
    m_P["Depressed", "Depressionfree"]<-  p_Depressed_Depressionfree
    m_P["Borderline", "Depressed"]<-p_Bordeline_Depressed
    m_P["Borderline", "Borderline"]<-p_Borderline_Borderline 
    m_P["Borderline", "Depressionfree"]<-  p_Borderline_Depressionfree
    m_P["Depressionfree", "Depressed"]<-p_Depressionfree_Depressed
    m_P["Depressionfree", "Borderline"]<-p_Depressionfree_Borderline
    m_P["Depressionfree", "Depressionfree"]<-p_Depressionfree_Depressionfree
    
    # State membership
    state_membership<-array(NA_real_,  dim= c(n_t, n_s), dimnames = list (cycle =1:n_t, state = v_state_names ))
    state_membership[1, ] <- c(n_c, 0, 0)
    for (i in 2:n_t) {state_membership[i, ] <-state_membership[i-1, ] %*% m_P}
    
    # payoffs
    m_payoffs <- matrix(0, nrow = 3, ncol = 2, 
                        dimnames = list(v_state_names, payoff= c("Cost", "HDS")))
    m_payoffs["Depressed", "Cost"]<- c_Depressed
    m_payoffs["Depressed", "HDS"]<- q_Depressed
    m_payoffs["Borderline", "Cost"]<- c_Borderline
    m_payoffs["Borderline", "HDS"]<- q_Borderline
    m_payoffs["Depressionfree", "Cost"]<- c_Depressionfree
    m_payoffs["Depressionfree", "HDS"]<- q_Depressionfree
    
    payoff_trace.se<-state_membership %*% m_payoffs
    payoff_trace.se
    
    summary_results.se = colSums(payoff_trace.se)/n_c 
    
  })
}
psa_results.se <-
  t(sapply(
    X = split(params, 1:N_psa), 
    FUN = model.se,
    simplify = TRUE ))

write.csv(psa_results.se, file="psa_coef3_se.csv")

plot(
  psa_results.se[, 2],
  psa_results.se[, 1],
  type = 'p',
  xlab = "HDS",
  ylab = "Cost"
)
```
Amitriptyline

Diagrammatic presentation of transition matrix
```{r ca}
Ami.matrix<-new("markovchain", states=c("Depressed", "Borderline", "Depression_free"), 
         transitionMatrix=matrix(data=c(0.2, 0.3, 0.5, 0.4, 0.2, 0.4, 0.2, 0.2, 0.6), 
                                   byrow = TRUE, nrow =  3), name = "Amitriptyline") 
plot(Ami.matrix, package = "daigram")
```
Markov cohort simulation with probability Sensitivity analysis of amitriptlyine

```{r c}
N_psa <- 10000

# R code inside a function

# distribution of transition probability matrix using beta distribution with SD= 25% of the mean value
params<-data.frame(
  p_Depressed_Depressed   = rbeta(N_psa, 20, 80),
  p_Depressed_Borderline  = rbeta(N_psa,  30, 70),
  p_Depressed_Depressionfree  = rbeta(N_psa,  50, 50),
  p_Bordeline_Depressed  = rbeta(N_psa,  40, 60),
  p_Borderline_Borderline = rbeta(N_psa, 20, 80),
  p_Borderline_Depressionfree = rbeta(N_psa,  40, 80),
  p_Depressionfree_Depressed  = rbeta(N_psa,  20, 80),
  p_Depressionfree_Borderline = rbeta(N_psa,  20,  80),
  p_Depressionfree_Depressionfree = rbeta(N_psa, 60, 40),
 
  # distribution of cost using gamma distribution with SD= 25% of the mean value 
  c_Depressed = rgamma(N_psa, shape = 16.05, scale = 41.19),
  c_Borderline = rgamma(N_psa, shape = 16, scale = 62), 
  c_Depressionfree = rgamma(N_psa, shape = 16.01, scale = 103.19),
  
  # distribution of cost using gamma distribution with SD= 25% of the mean value
  q_Depressed = rgamma(N_psa, shape = 16., scale = 0.125),
  q_Borderline = rgamma(N_psa, shape = 16.13, scale = 0.31),
  q_Depressionfree = rgamma(N_psa, shape = 16.05, scale = 0.5)
)

model.ami<- function(.params) { 
  with(.params, {
    
    n_t<-40
    n_s<-3
    n_c<-1000
    
    v_state_names<-c("Depressed", "Borderline", "Depressionfree")
    m_P<-matrix(0, nrow = 3, ncol =3, 
                dimnames = list(from = v_state_names, to = v_state_names ))
    m_P["Depressed", "Depressed"]<-p_Depressed_Depressed 
    m_P["Depressed", "Borderline"]<- p_Depressed_Borderline
    m_P["Depressed", "Depressionfree"]<-  p_Depressed_Depressionfree
    m_P["Borderline", "Depressed"]<-p_Bordeline_Depressed
    m_P["Borderline", "Borderline"]<-p_Borderline_Borderline 
    m_P["Borderline", "Depressionfree"]<-  p_Borderline_Depressionfree
    m_P["Depressionfree", "Depressed"]<-p_Depressionfree_Depressed
    m_P["Depressionfree", "Borderline"]<-p_Depressionfree_Borderline
    m_P["Depressionfree", "Depressionfree"]<-p_Depressionfree_Depressionfree
    
    # State membership
    state_membership<-array(NA_real_,  dim= c(n_t, n_s), dimnames = list (cycle =1:n_t, state = v_state_names ))
    state_membership[1, ] <- c(n_c, 0, 0)
    for (i in 2:n_t) {state_membership[i, ] <-state_membership[i-1, ] %*% m_P}
    
    # payoffs
    m_payoffs <- matrix(0, nrow = 3, ncol = 2, 
                        dimnames = list(v_state_names, payoff= c("Cost", "HDS")))
    m_payoffs["Depressed", "Cost"]<- c_Depressed
    m_payoffs["Depressed", "HDS"]<- q_Depressed
    m_payoffs["Borderline", "Cost"]<- c_Borderline
    m_payoffs["Borderline", "HDS"]<- q_Borderline
    m_payoffs["Depressionfree", "Cost"]<- c_Depressionfree
    m_payoffs["Depressionfree", "HDS"]<- q_Depressionfree
    
    payoff_trace.ami<-state_membership %*% m_payoffs
    payoff_trace.ami
    
    summary_results.ami = colSums(payoff_trace.ami)/n_c 
    
  })
}
psa_results.ami <-
  t(sapply(
    X = split(params, 1:N_psa), 
    FUN = model.ami,
    simplify = TRUE ))

write.csv(psa_results.ami, file="psa_coef3.csv")

plot(
  psa_results.ami[, 2],
  psa_results.ami[, 1],
  type = 'p',
  xlab = "HDS",
  ylab = "Cost"

)
```
Ecconomic evaluation

```{r d}
# incremental cost
data.result<- data.frame(psa_results.se, psa_results.ami)


inc_cost<- (data.result$Cost - data.result$Cost.1) 


# incremental effect
inc_effect<- (data.result$HDS - data.result$HDS.1)

# incremental cost effectiveness ratio
inc.CER <- inc_cost/inc_effect


# using a willingness to pay threshold of NGN18,000

wtp =  18000

psa_wtp <- sum(inc.CER <= wtp)

# percentage of cost effectiveness of sertraline over amitriptyline at wtp

result <- (psa_wtp /N_psa) * 100
 
result
```

session information

```{r e}
sessionInfo()
```

