---
title: "Cost-Effectiveness Analysis of Sertraline vs. Amitriptyline Using Markov Modelling"
author: "Joshua Edefo"
date: "2025-02-07"
oemail: "edefojoshua2000@yahoo.com"
output: github_document
---
This study employs a Markov cohort simulation and probabilistic sensitivity analysis (PSA) to compare the cost-effectiveness of Sertraline and Amitriptyline for depression. It models transitions, costs, and health benefits, calculating ICERs and CEACs to assess cost-effectiveness at different willingness-to-pay (WTP) thresholds. At a WTP threshold of NGN5400, Sertraline has a 78% probability of being more cost-effective than Amitriptyline.

Libraries
```{r message=FALSE}
library(cartography)
library(rmarkdown)
library(sf)
library(markovchain)
library(sp)
library(ggplot2)
library(dplyr)
library(usethis)
```

Sertraline

```{r b}
## Define parameters
Set.matrix<-new("markovchain", states=c("Depressed", "Borderline", "Depression free"), 
                transitionMatrix=matrix(data=c(0.1, 0.3, 0.6, 0.3, 0.1, 0.6, 0.1, 0.1, 0.8), 
                                        byrow = TRUE, nrow =  3), name = "Sertraline")
plot(Set.matrix, package = "daigram") 



## R code inside a function
N_psa <- 10000
## Monte Carlo simulations
## Distribution of transition probability matrix using beta distribution with SD= 25% of the mean value
params<-data.frame(
  p_Depressed_Depressed   = rbeta(N_psa, 0.1, 0.9),
  p_Depressed_Borderline  = rbeta(N_psa,  0.3, 0.7),
  p_Depressed_Depressionfree  = rbeta(N_psa,  0.6, 0.4),
  p_Bordeline_Depressed  = rbeta(N_psa,  0.3, 0.7),
  p_Borderline_Borderline = rbeta(N_psa, 0.1, 0.9),
  p_Borderline_Depressionfree = rbeta(N_psa,  0.6, 0.4),
  p_Depressionfree_Depressed  = rbeta(N_psa,  0.1, 0.9),
  p_Depressionfree_Borderline = rbeta(N_psa,  0.1,  0.9),
  p_Depressionfree_Depressionfree = rbeta(N_psa, 0.8, 0.2),
  
  ## distribution of cost using gamma distribution with SD= 25% of the mean value
  c_Depressed = rgamma(N_psa, shape =16, scale = 111/16),
  c_Borderline = rgamma(N_psa, shape =16, scale =62960/16),
  c_Depressionfree = rgamma(N_psa, shape =16, scale =95930/16),
  
  ## distribution of effect using gamma distribution with SD= 25% of the mean value
  q_Depressed = rgamma (N_psa, shape =16, scale = 3/16),
  q_Borderline = rgamma(N_psa, shape =16, scale = 12.5/16),
  q_Depressionfree = rgamma(N_psa, shape =16, scale = 19/16)
)

model.se<- function(.params) { 
  with(.params, {
    
    n_t<-6     ## no of cycles
    n_s<-3      ## no of states
    n_c<-500   ##  no of persons in cohort
    
    v_state_names<-c("Depressed", "Borderline", "Depressionfree") # names of states
    
    ## probability tranistion matrix
    m_P<-matrix(0, nrow = 3, ncol =3, 
                dimnames = list(from = v_state_names, to = v_state_names ))
    m_P["Depressed", "Depressed"]<-p_Depressed_Depressed 
    m_P["Depressed", "Borderline"]<- p_Depressed_Borderline
    m_P["Depressed", "Depressionfree"]<-  p_Depressed_Depressionfree
    m_P["Borderline", "Depressed"]<-p_Bordeline_Depressed
    m_P["Borderline", "Borderline"]<-p_Borderline_Borderline 
    m_P["Borderline", "Depressionfree"]<-  p_Borderline_Depressionfree
    m_P["Depressionfree", "Depressed"]<-p_Depressionfree_Depressed
    m_P["Depressionfree", "Borderline"]<-p_Depressionfree_Borderline
    m_P["Depressionfree", "Depressionfree"]<-p_Depressionfree_Depressionfree
    
    ## State membership
    state_membership<-array(NA_real_,  dim= c(n_t, n_s), dimnames = list (cycle =1:n_t, state = v_state_names ))
    state_membership[1, ] <- c(n_c, 0, 0)
    for (i in 2:n_t) {state_membership[i, ] <-state_membership[i-1, ] %*% m_P}
    
    ## payoffs
    m_payoffs <- matrix(0, nrow = 3, ncol = 2, 
                        dimnames = list(v_state_names, payoff= c("Cost", "HDS")))
    m_payoffs["Depressed", "Cost"]<- c_Depressed
    m_payoffs["Depressed", "HDS"]<- q_Depressed
    m_payoffs["Borderline", "Cost"]<- c_Borderline
    m_payoffs["Borderline", "HDS"]<- q_Borderline
    m_payoffs["Depressionfree", "Cost"]<- c_Depressionfree
    m_payoffs["Depressionfree", "HDS"]<- q_Depressionfree
    
    payoff_trace.se<-state_membership %*% m_payoffs
    payoff_trace.se
    
    summary_results.se = colSums(payoff_trace.se)/n_c 
    
  })
}
psa_results.se <-
  t(sapply(
    X = split(params, 1:N_psa), 
    FUN = model.se,
    simplify = TRUE ))

write.csv(psa_results.se, file="psa_coef3_se.csv")

plot(
  psa_results.se[, 2],
  psa_results.se[, 1],
  type = 'p',
  xlab = "HDS",
  ylab = "Cost"
)


```

Amitriptyline

```{r c}

## Defining parameters

Ami.matrix<-new("markovchain", states=c("Depressed", "Borderline", "Depression free"), 
                transitionMatrix=matrix(data=c(0.2, 0.4, 0.4, 0.4, 0.2, 0.4, 0.4, 0.4, 0.2), 
                                        byrow = TRUE, nrow =  3), name = "Amitriptyline")
plot(Ami.matrix, package = "daigram") 

N_psa <- 10000

## R code inside a function
## Monte Carlo simulations
## distribution of transition probability matrix using beta distribution with SD= 25% of the mean value
params<-data.frame(
  p_Depressed_Depressed   = rbeta(N_psa, 0.2, 0.8),
  p_Depressed_Borderline  = rbeta(N_psa,  0.4, 0.6),
  p_Depressed_Depressionfree  = rbeta(N_psa,  0.4, 0.6),
  p_Bordeline_Depressed  = rbeta(N_psa,  0.4, 0.6),
  p_Borderline_Borderline = rbeta(N_psa, 0.2, 0.8),
  p_Borderline_Depressionfree = rbeta(N_psa,  0.4, 0.6),
  p_Depressionfree_Depressed  = rbeta(N_psa,  0.4, 0.6),
  p_Depressionfree_Borderline = rbeta(N_psa,  0.4,  0.6),
  p_Depressionfree_Depressionfree = rbeta(N_psa, 0.2, 0.8),
  
  # distribution of cost using gamma distribution with SD= 25% of the mean value 
  c_Depressed = rgamma(N_psa, shape = 16, scale = 111/16),
  c_Borderline = rgamma(N_psa, shape = 16, scale = 24660/16), 
  c_Depressionfree = rgamma(N_psa, shape = 16.01, scale = 56770/16),
  
  # distribution of cost using gamma distribution with SD= 25% of the mean value
  q_Depressed = rgamma(N_psa, shape = 16, scale = 3/16),
  q_Borderline = rgamma(N_psa, shape = 16, scale = 10/16),
  q_Depressionfree = rgamma(N_psa, shape = 16, scale = 16/16)
)

model.ami<- function(.params) { 
  with(.params, {
    
    n_t<-40
    n_s<-3
    n_c<-1000
    
    v_state_names<-c("Depressed", "Borderline", "Depressionfree")
    m_P<-matrix(0, nrow = 3, ncol =3, 
                dimnames = list(from = v_state_names, to = v_state_names ))
    m_P["Depressed", "Depressed"]<-p_Depressed_Depressed 
    m_P["Depressed", "Borderline"]<- p_Depressed_Borderline
    m_P["Depressed", "Depressionfree"]<-  p_Depressed_Depressionfree
    m_P["Borderline", "Depressed"]<-p_Bordeline_Depressed
    m_P["Borderline", "Borderline"]<-p_Borderline_Borderline 
    m_P["Borderline", "Depressionfree"]<-  p_Borderline_Depressionfree
    m_P["Depressionfree", "Depressed"]<-p_Depressionfree_Depressed
    m_P["Depressionfree", "Borderline"]<-p_Depressionfree_Borderline
    m_P["Depressionfree", "Depressionfree"]<-p_Depressionfree_Depressionfree
    
    ## State membership
    state_membership<-array(NA_real_,  dim= c(n_t, n_s), dimnames = list (cycle =1:n_t, state = v_state_names ))
    state_membership[1, ] <- c(n_c, 0, 0)
    for (i in 2:n_t) {state_membership[i, ] <-state_membership[i-1, ] %*% m_P}
    
    ## payoffs
    m_payoffs <- matrix(0, nrow = 3, ncol = 2, 
                        dimnames = list(v_state_names, payoff= c("Cost", "HDS")))
    m_payoffs["Depressed", "Cost"]<- c_Depressed
    m_payoffs["Depressed", "HDS"]<- q_Depressed
    m_payoffs["Borderline", "Cost"]<- c_Borderline
    m_payoffs["Borderline", "HDS"]<- q_Borderline
    m_payoffs["Depressionfree", "Cost"]<- c_Depressionfree
    m_payoffs["Depressionfree", "HDS"]<- q_Depressionfree
    
    payoff_trace.ami<-state_membership %*% m_payoffs
    payoff_trace.ami
    
    summary_results.ami = colSums(payoff_trace.ami)/n_c 
    
  })
}
psa_results.ami <-
  t(sapply(
    X = split(params, 1:N_psa), 
    FUN = model.ami,
    simplify = TRUE ))

write.csv(psa_results.ami, file="psa_coef3.csv")

plot(
  psa_results.ami[, 2],
  psa_results.ami[, 1],
  type = 'p',
  xlab = "HDS",
  ylab = "Cost"
  
)


```

Ecconomic evaluation of Antidepressants

```{r d}
## incremental cost of sertraline over amitriptyline
data.result<- data.frame(psa_results.se, psa_results.ami)
data.result

inc_cost<- (data.result$Cost - data.result$Cost.1) 
inc_cost

## incremental effect of sertraline over amitriptyline
inc_effect<- (data.result$HDS - data.result$HDS.1)
inc_effect

## incremental cost effectiveness ratio
inc.CER <- inc_cost/inc_effect
inc.CER 

## using a willingness to pay threshold of NGN5400 

wtp =  5400

psa_wtp <- sum(inc.CER < wtp)

psa_wtp 

result <- (psa_wtp /N_psa) * 100

result

## Create a cost-effectiveness plane plot
icer_data<- data.frame(inc_cost,inc_effect)
icer_data
data.result
ggplot(icer_data, aes(x = inc_effect, y = inc_cost)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Incremental Effectiveness (HDS)",
    y = "Incremental Cost (NGN)",
    title = "Cost-Effectiveness Plane"
  ) +
  theme_minimal()

## Create an ICER curve (Cost-effectiveness acceptability curve)
## Define a range of WTP thresholds
wtp_thresholds <- seq(0, 16200, by = 300)

## Calculate the proportion of simulations where the ICER is below each WTP threshold
ceac <- sapply(wtp_thresholds, function(wtp) {
  mean(inc.CER  < wtp)
})

ceac_a <- 1 - ceac

## Create a data frame for the CEAC
ceac_data <- data.frame (
  wtp = wtp_thresholds,
  probability = ceac, ceac_a
)

## Plot the CEAC
ggplot(ceac_data, aes(x = wtp, y = probability)) +
  geom_line(color = "blue") +
  labs(
    x = "Willingness to Pay Threshold (NGN)",
    y = "Probability Cost-Effective",
    title = "Cost-Effectiveness Acceptability Curve"
  ) +
  theme_minimal()


## Create a data frame for the CEAC
ceac_data <- data.frame(
  wtp = wtp_thresholds,
  ceac = ceac,
  ceac_a = ceac_a
)

## Plot the CEAC and CEAC_A on the same graph
ggplot(ceac_data) +
  geom_line(aes(x = wtp, y = ceac, color = "Sertraline")) +
  geom_line(aes(x = wtp, y = ceac_a, color = "Amitrpyline")) +
  labs(
    x = "Willingness to Pay Threshold (NGN)",
    y = "Probability Cost-Effective",
    title = "Cost-Effectiveness Acceptability Curve of Sertraline versus Amitriptyline"
  ) +
  scale_color_manual(values = c("CEAC" = "blue", "CEAC_A" = "red")) +
  theme_minimal()


## Plot the CEAC and CEAC_A on the same graph
ggplot(ceac_data) +
  geom_line(aes(x = wtp, y = ceac, color = "Sertraline")) +
  geom_line(aes(x = wtp, y = ceac_a, color = "Amitriptyline")) +
  geom_text(aes(x = max(wtp), y = ceac[length(ceac)], label = "Sertraline"), 
            hjust = -0.2, color = "blue") +
  geom_text(aes(x = max(wtp), y = ceac_a[length(ceac_a)], label = "Amitriptyline"), 
            hjust = -0.2, color = "red") +
  labs(
    x = "Willingness to Pay Threshold (NGN)",
    y = "Probability Cost-Effective",
    title = "Cost-Effectiveness Acceptability Curve of Sertraline versus Amitriptyline"
  ) +
  scale_color_manual(values = c("Sertraline" = "blue", "Amitriptyline" = "red")) +
  theme_minimal()

```

session information

```{r e}
sessionInfo()
```